---
title: "Detecting and Correcting Overfitting in Meta-Regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detecting and Correcting Overfitting in Meta-Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(metaoverfit)
library(metafor)
```

## Introduction

In meta-regression, the $R^2_{het}$ statistic measures the proportion of between-study variance explained by moderators. However, when the number of studies ($k$) is small, the apparent $R^2_{het}$ can be highly optimistic due to overfitting. The `metaoverfit` package provides tools to estimate and correct for this optimism.

## Quick Start: Checking for Overfitting

The primary tool for diagnosing overfitting is the `check_overfitting()` function. It performs both cross-validation and bootstrap correction in one step.

```{r quick-start}
set.seed(42)

# Generate a dataset where moderators explain NO true heterogeneity (Null Model)
k <- 15 # Small number of studies
p <- 3  # Multiple parameters
yi <- rnorm(k, 0, sqrt(0.05)) # Random effects
vi <- runif(k, 0.01, 0.1)     # Sampling variances
mods <- cbind(1, rnorm(k), rnorm(k)) # Two random moderators + intercept

# Assess the risk of overfitting
result <- check_overfitting(yi, vi, mods, B = 500) # Use B=500+ for stable results
print(result)
```

In this example, even though the moderators were purely random noise, the **Apparent $R^2$** might be surprisingly high. The **Corrected $R^2$** correctly shrinks this towards zero.

## Understanding the Risk Categories

`metaoverfit` classifies the risk of overfitting based on the $k/p$ ratio and the observed optimism:

- **Low Risk:** $k/p \ge 15$. Results are generally stable.
- **Moderate Risk:** $k/p \approx 10-15$. Results should be reported with optimism correction.
- **Severe/Extreme Risk:** $k/p < 10$. Results are highly unreliable and may be entirely due to overfitting.

## Visualizing Overfitting

You can visualize the difference between apparent and corrected $R^2$ using `plot_overfitting()`.

```{r plotting, fig.width=7, fig.height=5}
# Bar plot comparison
plot_overfitting(result, type = "bar")

# If cross-validation (r2het_cv) was run, you can see observed vs predicted
cv_res <- r2het_cv(yi, vi, mods, verbose = FALSE)
plot_overfitting(cv_res, type = "scatter")

# Density plot of bootstrap distributions
plot_overfitting(result, type = "density")
```

## Planning for Future Studies: Sample Size Guidance

Before conducting a meta-regression, you can determine the minimum number of studies ($k$) required for a stable $R^2$ estimate using `sample_size_recommendation()`.

```{r sample-size}
# How many studies do I need for a model with 3 moderators (p=4)
# to keep optimism below 10%?
rec <- sample_size_recommendation(p = 4, target_optimism = 0.10)
print(rec)
```

## Using the Formula Interface

`metaoverfit` supports the standard R formula interface, making it compatible with `data.frame` objects.

```{r formula}
# Using the BCG vaccine dataset from metafor
data(dat.bcg, package = "metadat")
dat <- escalc(measure = "RR", ai = tpos, bi = tneg, 
               ci = cpos, di = cneg, data = dat.bcg)

# Check for overfitting in a model with ablat (latitude) and year
fit_overfit <- check_overfitting(yi, vi, mods = ~ ablat + year, 
                                 data = dat, B = 200)
print(fit_overfit)
```

## Conclusion

The `metaoverfit` package provides a necessary layer of protection against the over-interpretation of heterogeneity in meta-analysis. By reporting optimism-corrected $R^2$ values, researchers can provide more honest and reproducible evidence synthesis.
